---
title: "DF3-Citizenship"
author: "Abhay Prasanna Rao"
date: "2023-12-02"
output: html_document
---

```{r setup_citizenship, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

------------------------------------------------------------------------

## Citizenship


Citizenship status significantly influences employment prospects, with non-citizens often facing unique challenges in the job market. Legal restrictions and specific job requirements tied to citizenship can limit access to certain professions, hindering career opportunities and advancement for non-citizens. Employers may encounter administrative complexities when hiring individuals without citizenship, potentially impacting their hiring decisions. Discrimination based on immigration status can further exacerbate challenges for non-citizens in securing meaningful employment. Balancing national security considerations with the principles of equal employment opportunities is crucial to fostering inclusive workplaces that recognize and harness the skills and contributions of individuals, irrespective of their citizenship status. Efforts to promote fair employment policies and combat biases contribute to creating a more equitable job market for all.

```{r,message=FALSE,warning=FALSE}
#Esha
#Citizenship - K200501

df3<- get_acs(
  geography = "state", 
  table = "K200501", 
  year = 2021, 
  survey = "acs1/subject", 
  cache_table = TRUE
)

variables_df3 <- data.frame(
  variable = c("K200501_001", "K200501_002","K200501_003"),  
  label = c("Total","U.S. citizen", "Not a U.S. citizen")  
)

df3_labeled <- df3 %>%
  left_join(variables_df3, by = "variable")

df3_wide <- df3_labeled %>%
  pivot_wider(names_from = label, values_from = estimate, id_cols = NAME)

# Load necessary libraries
library(ggplot2)
library(tidyr)
library(dplyr)

# Convert df3_wide to long format
df3_long <- df3_wide %>%
  gather(key = "Citizenship", value = "Count", -NAME)

# Calculate total count for each state
df3_long <- df3_long %>%
  group_by(NAME) %>%
  mutate(Total = sum(Count)) %>%
  ungroup()

# Calculate the percentage
df3_long <- df3_long %>%
  mutate(Percentage = (Count / Total) * 100)

# Iterate over each unique citizenship category
for (category in unique(df3_long$Citizenship)) {
  # Filter data for the current citizenship category
  df3_subset <- df3_long %>% 
    filter(Citizenship == category)

  # Generate and print the plot for citizenship
  p <- ggplot(df3_subset, aes(x = NAME, y = Percentage, fill = Citizenship)) +
    geom_bar(stat = "identity") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(title = paste("Percentage Distribution of", category, "across States"), 
         x = "State", 
         y = "Percentage")

  print(p)
}




```
The ratio of US Citizens vs Non-US Citizens varies greatly across various states in the U.S. as you can observe from the graph above.In general there is a trend observed that , states with lesser US citizens have a lower employment rate.