---
title: "DF5-Housing"
author: "Abhay Prasanna Rao"
date: "2023-12-02"
output: html_document
---

```{r setup_housing, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```



------------------------------------------------------------------------

## Housing

Housing plays a pivotal role in influencing employment outcomes for individuals, as the accessibility, affordability, and stability of housing can significantly impact job opportunities and career advancement. Housing affordability directly affects an individual's financial well-being, and high housing costs relative to income may force individuals to allocate a substantial portion of their earnings to housing, leaving less for other essential needs. In some cases, individuals may face the challenge of commuting long distances due to housing affordability issues, which can impact their ability to secure and maintain employment. Moreover, stable and secure housing can contribute to a positive work environment, enabling individuals to focus on their careers without the stress and uncertainty associated with inadequate or unstable housing situations. Policies and initiatives that address housing affordability, promote stable living conditions, and encourage equitable urban development can play a crucial role in supporting individuals in their pursuit of meaningful and sustainable employment opportunities.

```{r,message=FALSE}
#Niharika
# Housing - K202502
df5 <- get_acs(geography = "state", 
                table = "K202502", 
                year = 2021, 
                survey = "acs1", # removed /subject since we're referring to a specific table ID
                cache_table = TRUE)

# Define the variable codes and their respective labels for the housing dataset (df5)
variables_df5 <- data.frame(
  variable = c("K202502_001", "K202502_002", "K202502_003"),
  label = c("Total", "Owner Occupied", "Renter Occupied")
)

# Assuming df5 is your housing dataset
# Reshape the data so that state names are rows
df5_labeled <- df5 %>%
  left_join(variables_df5, by = "variable")  # Joining with the variable labels dataframe

# Reshape the data to have state names as rows and variable labels as columns
df5_wide <- df5_labeled %>%
  pivot_wider(names_from = label, values_from = estimate, id_cols = NAME)

# df5_wide will have states as rows and different housing-related variables as columns


```

```{r,message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)

# Assuming df5_wide is your dataset for housing
# Melt the data to long format for easy plotting
df5_long <- tidyr::gather(df5_wide, key = "Housing_Category", value = "Count", -NAME)

# Calculate the percentage for Owner and Renter Occupied categories
df5_long <- df5_long %>%
  group_by(NAME) %>%
  mutate(Total = sum(Count[which(Housing_Category == "Total")])) %>%
  ungroup() %>%
  mutate(Percentage = ifelse(Housing_Category != "Total", Count / Total * 100, Count))

# Define colors for each category
category_colors <- c("Owner Occupied" = "steelblue", "Renter Occupied" = "salmon", "Total" = "lightgreen")

# Function to create a bar plot
create_bar_plot <- function(data, category, title) {
  ggplot(data[data$Housing_Category == category, ], aes(x = reorder(NAME, -Percentage), y = Percentage, fill = Housing_Category)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    labs(x = "State", y = "Percentage", title = title) +
    scale_y_continuous(labels = percent_format(scale = 1)) +
    scale_fill_manual(values = category_colors) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
}

# Plot for Owner Occupied
owner_plot <- create_bar_plot(df5_long, "Owner Occupied", "Percentage of Owner Occupied Housing by State")
print(owner_plot)

# Plot for Renter Occupied
renter_plot <- create_bar_plot(df5_long, "Renter Occupied", "Percentage of Renter Occupied Housing by State")
print(renter_plot)




```
The above two bar charts show's us the distrubution of precentage of renter and owner occupied housing by state. We can see that states like district of columbia, new york, california have the highest number of people who rent the houses. And, states like west virginia, maine, michigan have high number of people who own their own property.
