---
title: "Proposal Draft 1"
author: "Abhay Prasanna Rao"
date: "2023-11-12"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### \*\*Project Title:\*

```{r}
#install.packages("tidycensus")
library(tidycensus)
```

```{r}
#census_api_key(Sys.getenv("CENSUS_API_KEY"), install = TRUE)
```

```{r}
df1 <- get_acs(geography = "county", 
                table = "K200101", 
                year = 2021, 
                survey = "acs1/subject",cache_table = TRUE)
```

### **Project Idea Update:**

**Project Title**: Comparative Analysis of Population by Sex in Urban and Rural Areas Using ACS Data

#### Updated Description of the Data Set:

-   **Source**: 2022 American Community Survey (ACS) 1-Year Estimates.

-   **Focus**: Total population by sex in different geographic areas.

-   **Variables**: Population estimates by sex, geographic identifiers, and margin of error.

#### Adjusted Data Cleaning Steps:

1.  **Address Special Symbols**: Interpret and handle symbols like '-', 'N', '(X)', 'median-', 'median+', '**', and '**\*' as per the ACS guidelines.

2.  **Geographic Consistency**: Ensure alignment of geographic data with the OMB's 2020 delineations.

3.  **Handle Margin of Error**: Carefully interpret and use margin of error data, especially where it's larger than the estimate itself or treated as zero.

#### Updated Marginal Summaries:

-   **Population Breakdown**: Summarize population counts by sex across different regions.

-   **Error Analysis**: Evaluate the impact of margin of error on population estimates.

-   **Urban vs. Rural Comparison**: Initial comparison of population distributions in urban and rural settings based on sex.
