---
title: "Proposal Draft 1"
author: "Abhay Prasanna Rao"
date: "2023-11-12"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Project Title:**

**Comparative Analysis of Urban vs. Rural Living in the United States**

### **Objective:**

To examine and contrast the demographic, economic, and social conditions of urban and rural areas in the U.S., leveraging the extensive data available through the **`tidycensus`** package.

### **Data Source:**

U.S. Census and American Community Survey (ACS) data accessed via the **`tidycensus`** package in R.

### **Project Outline:**

#### Introduction

-   Overview of the urban-rural divide and its significance in societal development.

-   Objectives and potential impacts of the analysis.

#### Data Acquisition and Preprocessing

-   **Data Acquisition**: Instructions on retrieving data using **`tidycensus`**.

    ```{r}
   # install.packages("tidycensus")
    library(tidycensus)

    ```

-   API Key:

    ```{r}
   # census_api_key("your_api_key_here", install = TRUE)

    ```

-   Retrieve Data:\
```{r}
#data <- get_acs(geography = "county", 
#                variables = "variable_code", 
#                year = 2020, 
 #               survey = "acs5")

```

-   **Data Preprocessing**:

    -   **Cleaning**: Addressing missing or inconsistent data.

    -   **Categorization**: Segregating data based on urban and rural classifications.

    -   **Transformation**: Adapting data for meaningful analysis.

#### Exploratory Data Analysis (EDA)

-   Descriptive statistics to understand basic data characteristics.

-   Visualizations showcasing demographic, economic, and housing trends.

-   Initial identification of notable differences and similarities between urban and rural areas.

#### In-Depth Comparative Analysis

-   Examination of income levels, poverty rates, and employment trends.

-   Housing market dynamics: ownership, values, and types.

-   Educational attainment and its correlation with economic opportunities.

#### Geographic and Social Dynamics

-   Mapping and analysis of population distribution and migration trends.

-   Social variables analysis: family structures, languages spoken, and community engagement.

#### Conclusion

-   Summary of key findings and their implications.

-   Recommendations for policy and future research.

#### References

-   Comprehensive citation of data sources and other references.

### **Deliverables:**

-   Detailed report with analytical findings and visualizations.

-   Code repository for reproducibility and further exploration.

### **Tools and Technologies:**

-   R programming for data analysis.

-   **`tidycensus`** for data acquisition.

-   Packages like **`ggplot2`** for data visualization.

### **Skills Demonstrated:**

-   Data acquisition and preprocessing.

-   Statistical analysis and mapping.

-   Understanding of urban and rural socio-economic dynamics.