---
title: "DS 202 Final Project Report"
authors: ["Abhay", "Srika","Neha","Esha","Niharika"]
date: "2023-11-12"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

```


------------------------------------------------------------------------

# [Part 1]{.underline}

## Title

#### "Socio-Economic Factors Influencing Employment in the United States: A Comprehensive State-by-State Analysis"

## **Authors**

1.  Abhay Prasanna Rao

2.  Srika Raja

3.  Neha

4.  Esha

5.  Niharika

## **Abstract (TL;DR)**

This project investigates the impact of socio-economic factors on employment rates across U.S. states.
Utilizing ACS 2021 data, we explore relationships between employment and variables like education, citizenship, and housing.
Key findings include significant correlations that inform employment dynamics in the U.S.

## **Motivation**

We aim to analyze various socio-economic factors influencing employment in the U.S.
This study is crucial for understanding how different aspects like education, age, and housing contribute to employment rates, thereby aiding policymakers and researchers.

## Summary

We have imported data set from the ACS survey.
We have 6 child RMD files for this project which has the data analysis for the Employment, Education, Citizenship, Age, Housing, Disabilities Data Set (ACS 2021).\
Further, we started exploring each data set in detail and then we started combining each data set with the employment to see what results we can expect.
We did find many direct relationships with each data set on employment data set.
We have put our concluding results in the Final Report to help us stand by with our conclusions.

# [Part 2:]{.underline}

## Data Sets and Variables

The below data sets are from data.census.gov [ United States Census Bureau].
We shortlisted it based on ACS 2021, inclusive for all states in United States.

1.  [Employment - K202301]{.underline}

    | Variable              | Description                          |
    |-----------------------|--------------------------------------|
    | Total                 | Total Employment Data                |
    | In Labor Force        | Total People in Labor Force          |
    | Civilian labor force: | Total People in Civilian Labor Force |
    | Employed              | Total People Employed                |
    | Unemployed            | Total People Unemployed              |
    | In Armed Forces       | Total People in Armed Forces         |
    | Not in labor force    | Total People not in Labor Force      |

2.  [Education - K201501]{.underline}

    | Variable                               | Description                                                                  |
    |--------------------------|----------------------------------------------|
    | Education_Total_students               | Total Students in the Education Survery                                      |
    | Education_Below_9th grade              | Number of students who have completed 9th grade                              |
    | Education_9th to 12th grade_no diploma | Number of students who have completed 9th grade to 12th grade but no diploma |
    | Education_High_school_graduate         | Number of high school graduate students                                      |
    | Education_Some college_no degree       | Number of people enrolled into some college but have not acquired a degree   |
    | Education_Associates_degree            | Number of people with associates degree                                      |
    | Education_Bachelors_degree             | Number of people with bachelors degree                                       |
    | Education_Graduate_professional degree | Number of people with Graduate Degree                                        |

3.  [Citizenship - K200501]{.underline}

    | Variable           | Description                              |
    |--------------------|------------------------------------------|
    | Total              | Total Number of people in survey         |
    | U.S. citizen       | Number of US citizen in the survery      |
    | Not a U.S. citizen | Number of Non US Citizens in the survery |

4.  [Age - K200104]{.underline}

    | Variable      | Description                                  |
    |---------------|----------------------------------------------|
    | Total_age     | Total number of people in the age data frame |
    | Age_under_18  | Total number of under 18 people              |
    | "Age_18_to_24 | People between 18 to 24                      |
    | Age_25_to_34  | People between 25 to 34                      |
    | Age_35_to_44  | People between 35 to 44                      |
    | Age_45_to_54  | People between 45 to 54                      |
    | Age_55_to_64  | People between 55 to 64                      |
    | Age_over_64   | People over 64                               |

5.  [Housing - K202502]{.underline}

    | Variable        | Description                                    |
    |-----------------|------------------------------------------------|
    | Total           | Total Number of People in housing data frame   |
    | Owner Occupied  | Total Number of people who have their own home |
    | Renter Occupied | Total Number of people who are renting a place |

6.  [Disabilities - K201803]{.underline}

    | Variable                | Description                     |
    |-------------------------|---------------------------------|
    | Total_people            | Total Poeple in the data frame  |
    | Total With Disabilities | Total with disabilities         |
    | Hearing                 | Total with hearing problem      |
    | Vision difficulty       | Total with vision problem       |
    | cognative               | Total with cognative problem    |
    | ambulatory difficulty   | Total with ambulatory difficlty |
    | Self-care difficulty    | Total with self care difficulty |
    | No Disability           | Total without disabilities      |

```{r, child=c('DS202_Final_Project.Rmd')}
```

# [Part 3]{.underline}

## Results

1.  Employment vs Education

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# Reshaping data: Education levels need to be in long format
long_data_education <- merged_data %>%
  select(NAME, Employed,`Education_Below_9th grade`) %>%
  pivot_longer(cols = c(`Education_Below_9th grade`), names_to = "EducationLevel", values_to = "EducationValue") %>%
  arrange(NAME)

# Now create the plot
ggplot(long_data_education, aes(x = EducationValue, y = Employed, color = EducationLevel)) +
  geom_point() +
  labs(title = "Employment vs Education Levels",
       x = "Number of Individuals with Education Level",
       y = "Number of Employed Individuals",
       color = "Education Level") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis labels for better readability


```

2.  Employment Vs Housing

```{r}
# Reshaping data: 'Employed', 'Owner Occupied', and 'Renter Occupied' need to be in long format
long_data <- merged_data %>%
  select(NAME, Employed, `Owner Occupied`, `Renter Occupied`) %>%
  pivot_longer(cols = c(`Owner Occupied`, `Renter Occupied`), names_to = "HousingType", values_to = "HousingValue") %>%
  arrange(NAME)

# Now create the grouped bar chart
ggplot(long_data, aes(x = NAME, y = Employed, fill = HousingType)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Employed Individuals vs States by Housing Type",
       x = "State",
       y = "Number of Employed Individuals",
       fill = "Housing Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis labels for better readability

```

3.  Citizenship Vs Employmennt

```{r}

# Reshaping data: 'Employed', 'U.S. citizen', and 'Not a U.S. citizen' need to be in long format
long_data_citizenship <- merged_data %>%
  select(NAME, Employed, `U.S. citizen`, `Not a U.S. citizen`) %>%
  pivot_longer(cols = c(`U.S. citizen`, `Not a U.S. citizen`), names_to = "CitizenshipStatus", values_to = "CitizenshipValue") %>%
  arrange(NAME)

# Now create the grouped bar chart
ggplot(long_data_citizenship, aes(x = NAME, y = Employed, fill = CitizenshipStatus)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Employed Individuals vs States by Citizenship Status",
       x = "State",
       y = "Number of Employed Individuals",
       fill = "Citizenship Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis labels for better readability

```

## **Conclusions**

This study provides insights into the socio-economic factors affecting employment in the U.S.
Limitations include data scope and potential biases.
Future work could explore more granular data and additional variables.

# [Part 4]{.underline}

## **Data Source**

1.  Data from the American Community Survey (ACS) 2021.
2.  Scripts for data cleaning and analysis are included in the repository.
3.  6 Child RMD files included in the repo, which support the DS202_final_project.rmd file

## **References**

-   American Community Survey (ACS)

-   United States Census Bureau (data.census.gov)
