---
title: "DS 202 Final Project Report"
authors: ["Abhay", "Srika","Neha","Esha","Niharika"]
date: "2023-11-12"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r, child=c('DF1-Employment.Rmd','DF2-Education.Rmd','DF3-Citizenship.Rmd','DF4-Age.Rmd','DF5-Housing.Rmd','DF6-Disabilities.Rmd')}
```

------------------------------------------------------------------------

# [Part 1]{.underline}

## Title

#### "Socio-Economic Factors Influencing Employment in the United States: A Comprehensive State-by-State Analysis"

## **Authors**

1.  Abhay Prasanna Rao

2.  Srika Raja

3.  Neha

4.  Esha

5.  Niharika

## **Abstract (TL;DR)**

This project investigates the impact of socio-economic factors on employment rates across U.S. states.
Utilizing ACS 2021 data, we explore relationships between employment and variables like education, citizenship, and housing.
Key findings include significant correlations that inform employment dynamics in the U.S.

## **Motivation**

We aim to analyze various socio-economic factors influencing employment in the U.S.
This study is crucial for understanding how different aspects like education, age, and housing contribute to employment rates, thereby aiding policymakers and researchers.

## Summary

We have imported data set from the ACS survey.
We have 6 child RMD files for this project which has the data analysis for the Employment, Education, Citizenship, Age, Housing, Disabilities Data Set (ACS 2021).\
Further, we started exploring each data set in detail and then we started combining each data set with the employment to see what results we can expect.
We did find many direct relationships with each data set on employment data set.
We have put our concluding results in the Final Report to help us stand by with our conclusions.

# [Part 2:]{.underline}

## Data Sets

The below data sets are from data.census.gov [ United States Census Bureau].
We shortlisted it based on ACS 2021, inclusive for all states in United States.

1.  Employment - K202301

    | Variable | Description | Range |
    |----------|-------------|-------|
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |

2.  Education - K201501

    | Variable | Description | Range |
    |----------|-------------|-------|
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |

3.  Citizenship - K200501

    | Variable | Description | Range |
    |----------|-------------|-------|
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |

4.  Age - K200104

    | Variable | Description | Range |
    |----------|-------------|-------|
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |

5.  Housing - K202502

    | Variable | Description | Range |
    |----------|-------------|-------|
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |

6.  Disabilities - K201803

    | Variable | Description | Range |
    |----------|-------------|-------|
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |
    |          |             |       |

## 
